BibleGoAI is composed of three major subsystems: (1) a data layer housing the biblical text and commentary corpus, (2) an AI pipeline that generates, scores, and validates BHTs, and (3) a web application that serves the results to end users. Figure~\ref{fig:architecture} provides an overview of the end-to-end architecture.

% --- ARCHITECTURE DIAGRAM ---
\begin{figure}[t]
\centering
\begin{tikzpicture}[
    node distance=0.5cm,
    block/.style={rectangle, draw=pipeline!40, fill=pipeline!8, text width=4.5cm, minimum height=0.7cm, align=center, font=\scriptsize\sffamily, rounded corners=2pt, line width=0.4pt},
    data/.style={rectangle, draw=tier1!40, fill=tier1!8, text width=4.5cm, minimum height=0.7cm, align=center, font=\scriptsize\sffamily, rounded corners=2pt, line width=0.4pt},
    output/.style={rectangle, draw=accent!40, fill=accent!8, text width=4.5cm, minimum height=0.7cm, align=center, font=\scriptsize\sffamily, rounded corners=2pt, line width=0.4pt},
    arrow/.style={-{Stealth[length=2mm]}, line width=0.6pt, color=pipeline!70},
    grouplabel/.style={font=\tiny\sffamily\bfseries, anchor=east}
]
% Data sources
\node[data] (data) {NASB Bible (66 books) + Commentary Corpus (14)};

% Pipeline steps - straight vertical
\node[block, below=0.8cm of data] (extract) {1. Quote Extraction \quad {\tiny\itshape GPT-3.5}};
\node[block, below=of extract] (retrieve) {2. Retrieval \quad {\tiny\itshape FAISS + HuggingFace}};
\node[block, below=of retrieve] (generate) {3. BHT Generation \quad {\tiny\itshape GPT-3.5}};
\node[block, below=of generate] (score) {4. Scoring \& Validation \quad {\tiny\itshape USE + textstat}};
\node[block, below=of score] (footnote) {5. Footnoting + Tagging \quad {\tiny\itshape Cross-encoder}};

% Output
\node[output, below=0.8cm of footnote] (deploy) {PostgreSQL $\rightarrow$ BibleGo.org (Next.js)};

% Arrows
\draw[arrow] (data) -- (extract);
\draw[arrow] (extract) -- (retrieve);
\draw[arrow] (retrieve) -- (generate);
\draw[arrow] (generate) -- (score);
\draw[arrow] (score) -- (footnote);
\draw[arrow] (footnote) -- (deploy);

% Background groups
\begin{scope}[on background layer]
    \node[fit=(data), fill=tier1!5, draw=none, rounded corners=4pt, inner sep=5pt] {};
    \node[fit=(extract)(retrieve)(generate)(score)(footnote), fill=pipeline!5, draw=none, rounded corners=4pt, inner sep=8pt] {};
    \node[fit=(deploy), fill=accent!5, draw=none, rounded corners=4pt, inner sep=5pt] {};
\end{scope}

% Group labels
\node[grouplabel, tier1] at ([xshift=-0.3cm]data.west) {Data};
\node[grouplabel, pipeline] at ([xshift=-0.3cm]generate.west) {Pipeline};
\node[grouplabel, accent] at ([xshift=-0.3cm]deploy.west) {Deploy};
\end{tikzpicture}
\caption{End-to-end system architecture. Commentary and Bible text flow through the five-phase AI pipeline, producing scored BHTs with footnotes that are stored in PostgreSQL and served via BibleGo.org.}
\label{fig:architecture}
\end{figure}

\subsection{Data Sources}

The system operates on two primary data sources:

\paragraph{Biblical text.} The New American Standard Bible (NASB, 1995 edition) provides the target verse text for all 66 books of the Bible. Each verse is stored as a JSON record with chapter, verse number, and text content. The NASB was selected for its reputation as a highly literal translation, which aligns well with the commentary corpus.

\paragraph{Commentary corpus.} The commentary collection comprises 14 curated commentators spanning the 16th through 20th centuries (four centuries of biblical scholarship). For the New Testament, nine commentators form the active corpus: Henry Alford, Jamieson-Fausset-Brown, Albert Barnes, Marvin Vincent, John Calvin, Philip Schaff, Archibald T. Robertson, John Gill, and John Wesley. They are organized into three tiers (Table~\ref{tab:nt-commentators}). For the Old Testament, twelve commentators are used: Jamieson-Fausset-Brown, Albert Barnes, John Gill, Keil \& Delitzsch, Cyrus I. Scofield, Johann P. Lange, John Calvin, William Kelly, Charles H. Mackintosh, Charles A. Coates, Arthur W. Pink, and Henri L. Rossier, similarly tiered (Table~\ref{tab:ot-commentators}). Four commentators (Jamieson-Fausset-Brown, Albert Barnes, John Gill, John Calvin) appear in both the NT and OT corpora, sometimes at different tiers.

All commentaries were curated and vetted by a group of Christians from the Cleveland, Ohio area, ensuring theological reliability and scholarly quality.

\subsection{Commentator Tiering}

The tiering system serves a critical role in the generation process. When synthesizing a BHT, the LLM is instructed to draw \textit{primarily} from Tier~1 commentators, \textit{sparingly} from Tier~2, and \textit{minimally} from Tier~3. This weighted approach ensures that the most trusted and comprehensive commentators dominate the output while allowing supplementary perspectives to enrich the synthesis. Tables~\ref{tab:nt-commentators} and~\ref{tab:ot-commentators} list the active NT and OT commentators with their tier assignments and approximate date ranges.

\begin{table}[htbp]
\centering
\small
\caption{Active New Testament commentators and tier assignments. Tier assignments are defined per testament; several commentators (Jamieson-Fausset-Brown, Albert Barnes, John Gill, John Calvin) appear in both the NT and OT corpora and may receive different tiers.}
\label{tab:nt-commentators}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Commentator} & \textbf{Era} & \textbf{Tier} \\
\midrule
Henry Alford & 1810--1871 & 1 \\
Jamieson, Fausset \& Brown & 1871 & 1 \\
Marvin R. Vincent & 1834--1922 & 1 \\
Archibald T. Robertson & 1863--1934 & 1 \\
Albert Barnes & 1798--1870 & 2 \\
Philip Schaff & 1819--1893 & 2 \\
John Wesley & 1703--1791 & 3 \\
John Gill & 1697--1771 & 3 \\
John Calvin & 1509--1564 & 3 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\small
\caption{Active Old Testament commentators and tier assignments.}
\label{tab:ot-commentators}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Commentator} & \textbf{Era} & \textbf{Tier} \\
\midrule
Keil \& Delitzsch & 1807--1890 & 1 \\
Arthur W. Pink & 1886--1952 & 1 \\
Henri L. Rossier & 1835--1928 & 1 \\
Charles A. Coates & 1862--1945 & 1 \\
Charles H. Mackintosh & 1820--1896 & 1 \\
Jamieson, Fausset \& Brown & 1871 & 2 \\
Johann P. Lange & 1802--1884 & 2 \\
William Kelly & 1821--1906 & 2 \\
Cyrus I. Scofield & 1843--1921 & 2 \\
Albert Barnes & 1798--1870 & 3 \\
John Gill & 1697--1771 & 3 \\
John Calvin & 1509--1564 & 3 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Prompt Engineering}

The pipeline relies on three families of prompts, each iteratively refined through multiple versions and recorded in the project repository: \textbf{choicest} (per-commentator quote extraction), \textbf{BHT} (single-paragraph synthesis, with both a single-pass \emph{regular} variant and a two-stage \emph{split} variant), and \textbf{tag commentary} (verse-level indexing of raw commentary text used during preprocessing). A fourth family, \emph{choicest range}, was developed to extract single-verse quotes from multi-verse commentary but did not reach the production pipeline; verse-range handling was instead addressed through retrieval-augmented generation (Section~\ref{sec:methods}, Phase~2) combined with the split pipeline for narrative OT books. The full prompt specifications, including the word-count and quote-proportion constraints used in production, appear in Section~\ref{sec:methods}, Phases~1 and~3.

\subsection{Commentary Digitization}

Not all commentaries in the corpus were available in digital form. Two major digitization efforts were undertaken by volunteers to expand the commentary corpus:

\paragraph{Alford's New Testament for English Readers.} Henry Alford's four-volume commentary (1868--1872) was digitized from physical book scans totaling 2{,}544 pages across all 27 New Testament books. The pipeline consisted of three stages: (1)~\textit{extraction} via Tesseract OCR and Google Cloud Vision, converting scanned pages to text; (2)~\textit{human correction}, in which 28 volunteers spent hundreds of hours reviewing and correcting OCR errors through a coordinated Google Drive spreadsheet; and (3)~\textit{automated consolidation}, which detected chapter boundaries, parsed verse-level commentary, performed spell-checking, and produced structured output files. This effort represents a significant scholarly contribution independent of the AI pipeline.

% --- DIGITIZATION PIPELINE DIAGRAM ---
\begin{figure}[t]
\centering
\begin{tikzpicture}[
    node distance=0.6cm,
    digbox/.style={rectangle, draw=tier1!40, fill=tier1!8, text width=2.8cm, minimum height=0.8cm, align=center, font=\scriptsize\sffamily, rounded corners=2pt, line width=0.4pt},
    digarrow/.style={-{Stealth[length=2mm]}, line width=0.5pt, color=tier1!60},
]
\node[digbox] (scan) {Book Scans\\(2{,}544 pages)};
\node[digbox, below=of scan] (ocr) {Tesseract OCR +\\Google Cloud Vision};
\node[digbox, below=of ocr] (human) {28 Volunteers\\(Google Drive)};
\node[digbox, below=of human] (consol) {Automated\\Consolidation};

\draw[digarrow] (scan) -- (ocr);
\draw[digarrow] (ocr) -- (human);
\draw[digarrow] (human) -- (consol);

\node[font=\tiny\sffamily\itshape, color=tier1!60, right=0.15cm of scan] {PDF/images};
\node[font=\tiny\sffamily\itshape, color=tier1!60, right=0.15cm of ocr] {Raw DOCX};
\node[font=\tiny\sffamily\itshape, color=tier1!60, right=0.15cm of human] {Corrected text};
\node[font=\tiny\sffamily\itshape, color=tier1!60, right=0.15cm of consol] {Structured JSON};
\end{tikzpicture}
\caption{Commentary digitization pipeline used for Alford's and Govett's commentaries.}
\label{fig:digitization}
\end{figure}

\paragraph{Govett's Commentary on Revelation.} Robert Govett's commentary on Revelation was digitized through the same volunteer-driven process, spanning 627 pages and producing 274 individual verse commentary files covering all 22 chapters. Both digitization projects followed the same three-stage pipeline, demonstrating its replicability for future commentary digitization.

\subsection{Original Language Tagging}

A separate pipeline was developed to tag every word in the Bible with its corresponding Hebrew or Greek Strong's concordance number. The raw NASB text, with original-language tags provided courtesy of the NASB, was processed through a dedicated parser that produced two outputs:

\begin{itemize}[nosep,leftmargin=*]
    \item \textbf{Bible tagged files}: 66 JSON files (one per book), where each English word is linked to its Strong's code (e.g., ``beginning'' $\rightarrow$ H7225).
    \item \textbf{Tagged occurrence index}: 14{,}141 individual Strong's code files containing all 321{,}621 verse occurrences across the Bible, enabling reverse lookup from any Hebrew or Greek word to every verse where it appears.
\end{itemize}

\noindent A dictionary of 9{,}333 Hebrew and 5{,}564 Greek entries was also compiled, providing original-language forms, transliterations, and definitions. This tagging infrastructure powers the Word Donut feature on BibleGo.org (Section~\ref{sec:platform}) and also enabled the TAG retrieval strategy for OT BHT generation (Section~\ref{sec:methods}).
