The study of the Bible has generated one of the largest and most enduring bodies of scholarly commentary in human history. Prominent theologians, from John Calvin in the sixteenth century to Archibald T. Robertson in the twentieth, have produced detailed verse-by-verse analyses that illuminate the linguistic, historical, and theological dimensions of the biblical text. For many Bible studiers, these commentaries are indispensable. Yet the sheer volume of material presents a practical challenge: consulting even a handful of commentators for a single verse requires navigating multiple lengthy texts, synthesizing sometimes divergent perspectives, and extracting the most relevant insights.

BibleGoAI grew out of a simple goal: \textit{help people study the Bible using time-tested commentaries presented in a simple and accessible way}, not only for theological understanding but also as a help to believers in their spiritual growth. The original commentary texts were sourced from StudyLight.org.

Modern advances in natural language processing (NLP) and large language models (LLMs) offer new possibilities for addressing this challenge. Systems based on retrieval-augmented generation (RAG) have demonstrated the ability to ground LLM outputs in authoritative source documents \citep{lewis2020rag}, while sentence embeddings \citep{cer2018use, reimers2019sentencebert} enable fine-grained semantic comparison between generated and source texts. These capabilities make it feasible to build systems that not only summarize but also \textit{validate} the fidelity of their outputs against the original material.

In this paper, we present \textbf{BibleGoAI}, a multi-stage pipeline that transforms curated biblical commentary into concise, source-adherent summary paragraphs called \textit{Brief Helpful Texts} (BHTs). The name reflects the project's core value: bringing the richness of difficult-to-read commentary to the fingertips of Bible readers in an accessible way. The system is designed around three principles:

\begin{enumerate}[nosep,leftmargin=*]
    \item \textbf{Fidelity}: BHTs must be grounded in the actual words of trusted commentators, not in the LLM's own knowledge. The pipeline enforces this with a hard gate (20--90\% of each BHT's words must originate from source quotes) and a tighter prompt-level target (30--90\% with a 70\% target stated to the model); the realized corpus-wide proportions are reported in Section~\ref{sec:evaluation}.
    \item \textbf{Conciseness}: Each BHT is a short paragraph of roughly 80 words and a few sentences (typically 2--4; see Section~\ref{sec:evaluation} for per-pipeline distributions), making it practical for quick reference during Bible study.
    \item \textbf{Traceability}: Every claim in a BHT is linked back to its source commentary via footnotes with similarity scores, enabling readers to verify and explore further.
\end{enumerate}

The pipeline was run once over the entire New Testament (7{,}957 verses across 27 books) and the full Old Testament (all 39 books), using two pipeline variants: a single-pass regular pipeline for 24 books and a two-stage split pipeline for 20 books (see Section~\ref{sec:evaluation} for details). It draws from a curated corpus of 17 unique commentators (nine NT, twelve OT, four shared) spanning four centuries and organized into three tiers of scholarly importance. The NT was publicly launched at a New Year's church conference in December 2023, with the OT following at a Memorial Day church conference in 2024. The resulting BHTs and their source commentaries are served to end users through \textbf{BibleGo.org}, a web application that provides interactive commentary exploration, original-language word study via Strong's concordance, and full commentator browsing, serving approximately 100--300 unique users per month with monthly page-view volumes ranging from ${\sim}200$ to ${\sim}2{,}500$ clicks depending on seasonal and community activity.

Our contributions are as follows:
\begin{itemize}[nosep,leftmargin=*]
    \item A multi-stage pipeline for commentary synthesis combining LLM-based quote extraction, tiered prompt engineering, and constraint-enforced generation.
    \item A composite quality scoring framework that integrates semantic similarity and source fidelity checks.
    \item An OT-specific two-pipeline design (a single-pass regular pipeline and a two-stage split pipeline) that together covers all 39 Old Testament books, with per-pipeline metrics reported across the generated corpus.
    \item A user-facing application (BibleGo.org) serving the generated BHTs to a live Bible-study community, demonstrating the practical impact of the system.
\end{itemize}

\paragraph{Project context.} BibleGoAI is a volunteer-driven collaboration among Christians, mostly based in the Cleveland, Ohio area, with backgrounds in software engineering, machine learning, and pastoral ministry. The project proceeded through weekly round-table meetings in 2023, a volunteer commentary digitization effort (Henry Alford's 2{,}544-page \textit{New Testament for English Readers} commentary, digitized by 28 volunteers in 2023, and Robert Govett's 627-page \textit{Commentary on Revelation}, digitized through the same process and completed in 2025), and an intensive April--May 2024 development sprint. The NT was publicly launched at a New Year's church conference in December 2023 and the OT at a Memorial Day church conference in 2024. Several technical decisions were shaped by this community context: the commentator corpus was curated by ministry leaders for theological reliability; usability testing was conducted with members of the project's Bible study community; the NASB~1995 translation was adopted after a request to use the ESV was declined by its publisher's licensing terms; and a sustained cadence of internal demos served as the primary feedback mechanism. These community and process constraints are a meaningful part of what this paper documents.
